Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2016-01-14 docs/appendixB.Rmd
update appB

Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig. B1). Thus, we set the kernal bandwidth ($\sigma$ in Eq. X of the main text) to 500/3 = `r round(500/3,2)`. We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters. Knots were placed evenly across space in the x and y directions. This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig. B2).

>>>>

Spatial dependence existed in the residuals until pixels were separated by about 500 meters (Fig. B1). Thus, we set the kernel bandwidth ($\sigma$ in Eq. 6 of the main text) to 500/3 = `r round(500/3,2)` because the effective spatial range under an exponential covariance structure is equal to $\sigma \times 3$. We then placed equally-spaced knots across our spatial grid until the distance between nearest neighbors was approximately 500 meters. Knots were placed evenly across space in the x and y directions. This resulted in a 11 $\times$ 21 knot grid (231 knots; Fig. B2).

### 2016-01-14 docs/ms_sageAbundance.Rmd
adds mention of shared climate variables earlier
update appB

Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data. Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; Xian2012; @Harte2015; Homer2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. 
Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. 
In warmer parts of its range, increased temperature could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. 
In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assessing how sagebrush could respond to climate change in the future.

@@

### Sagebrush response to climate and climate change
The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*, Fig. 3).
The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, mean estimates for the climate effects are relatively weak (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
We used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability. 
Nonetheless, our model was capabable of detecting climate effects that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.

>>>>

Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data. Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Xian2012; @Harte2015; @Homer2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. 
Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. 
In warmer parts of its range, increased temperature could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. 
In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assessing how sagebrush could respond to climate change in the future.

@@

### Sagebrush response to climate and climate change
The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*, Fig. 3).
The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, mean estimates for the climate effects are relatively weak (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
We used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability.
Additionally, we chose not to downscale the Daymet weather data, meaning that in a given year all pixels shared the same climate, which limits our statistical power.
Nonetheless, our model was capabable of detecting climate effects that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
